### 2023-05-09 chapter2/chapter2.tex
Added reference to CV figures for matrices
Converted substitution outline into paragraph

To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}, beginning with the frequency vectors. The disorder regions show an enrichment of ``disorder-promoting'' residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A). The exchangeability matrices fit to the disorder and order regions have similar overall patterns of high and low coefficients (Fig.~\ref{fig:substitution}B-C). However, the log ratios of the disorder to the order exchangeability coefficients show clear differences within and between the disorder-enriched and -depleted residues. In disorder regions, disorder-enriched residues are less exchangeable with each other, whereas disorder-depleted residues are more exchangeable with each other and with disorder-enriched residues (Fig.~\ref{fig:substitution}D). Likewise, we observe a trend in the log ratios of the rate coefficients where the coefficients above the diagonal are generally positive and those below the diagonal are generally negative. As the coefficients model the one-way rate of substitutions between residues with the vertical and horizontal axes indicating the initial and target residues, respectively, this suggests a net flux towards a more disorder-like composition.

@@

% II. IDRs have characteristic patterns of amino acid substitution
%    A. In addition to these global patterns of conserved features, we find IDRs exhibit other characteristic patterns of evolution.
%    B. For example, a comparison of the exchangeability matrices fit to the disorder and order regions show decreased exchangeability coefficients between
%    A. Interestingly, the exchangeability matrices suggest that although there is net flux towards disorder by depleting structure-promoting residues, disordered residues are more biochemically distinct in disordered regions than in structured regions. This suggests that although disorder-promoting amino acids are often grouped into a single category and described as largely hydrophilic, they have distinct roles in disordered regions. There is evidence that highly similar residues such as glutamine and asparagine have distinct properties in disordered regions that yielded different biophysical characters of phase-separated condensates, so this aligns with other observations. Conversely, structure promoting residues are more interchangeable. One hypothesis is as these bulkier hydrophobic residues are not constrained by packing, they are free to act more generically as small hydrophobic patches that mediate interactions in SLiMs and other binding regions in IDRs. Another model for this behavior is the stickers and spacers model of phase separation.
%    A. A caveat of this analysis is its dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models. Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences are re-aligned with scoring matrices derived from those models in alternating rounds until convergence. Because our analyses fit the substitution models directly from alignments, the observed patterns we observe could be artifacts of using a scoring matrix largely fit to folded domains to align IDRs
%    A. However, as the exchangeability and rate matrices broadly reproduced the trends observed in previous studies which fit amino acid substitution models to disorder regions, these observations are likely not artifactual but instead represent reproducible differences in the substitution rates between residues.
%    B. Interpretations of exchangeability and rate matrices
%        1. Disorder-enriched residues are less interchangeable and play specific roles in these regions
%        2. Disorder-depeleted residues not constrained by packing and can act more generically as small hydrophobic patches
%            i. Stickers and spacers from Holehouse
%            ii. Staller papers and other high throughput mutational scanning of ADs
%            iii. Halfmann glutamine vs asparagine paper
%        3. Note about variances of parameters as shown in CV graphs

>>>>

To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}, beginning with the frequency vectors. The disorder regions show an enrichment of ``disorder-promoting'' residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A). The exchangeability matrices fit to the disorder and order regions have similar overall patterns of high and low coefficients (Fig.~\ref{fig:substitution}B-C). However, the log ratios of the disorder to the order exchangeability coefficients show clear differences within and between the disorder-enriched and -depleted residues. In disorder regions, disorder-enriched residues are less exchangeable with each other, whereas disorder-depleted residues are more exchangeable with each other and with disorder-enriched residues (Fig.~\ref{fig:substitution}D). Likewise, we observe a trend in the log ratios of the rate coefficients where the coefficients above the diagonal are generally positive and those below the diagonal are generally negative. As the coefficients model the one-way rate of substitutions between residues with the vertical and horizontal axes indicating the initial and target residues, respectively, this suggests a net flux towards a more disorder-like composition. However, the coefficients between the disorder-depleted and -enriched classes of residues for both the exchangeability and rate matrices should be interpreted with caution, as they are estimated with a high amount of uncertainty (Fig.~\ref{sfig:CV_ematrix}-\ref{sfig:CV_rmatrix}).

@@

\subsection{IDRs have distinct patterns of sequence and feature evolution}
In addition to these global clusters of conserved features, we find IDRs are characterized by other distinct patterns of evolution. For example, a comparison of the exchangeability matrices fit to the disorder and order regions shows that, relative to the order regions, the disorder regions have decreased exchangeability coefficients between the disorder-enriched residues. Conversely, the disorder-depleted residues have increased exchangeability coefficients with each other. As a residue's enrichment in IDRs is generally interpreted as a measure of its ability to promote intrinsic disorder, these results indicate that within IDRs and relative to folded domains, disorder-promoting residues are subject to stricter constraints whereas structure-promoting residues are more biochemically interchangeable. A potential weakness of this analysis is its dependence on sequence alignments which were created using scoring matrices that are in turn derived from other substitution models. Previous studies have attempted to minimize the impact of this circular dependency through an EM-like procedure where substitution models are first fit to alignments, and the sequences in the alignments are then re-aligned with scoring matrices derived from the fit models in alternating rounds until convergence~\cite{Brown2009, Szalkowski2011}. Because our analyses fit the substitution models directly from alignments, the observed patterns are possibly an artifact of using a scoring matrix derived from folded domains to align IDRs. However, as our matrices reproduce the trends reported in these prior studies, they likely reflect true differences in the patterns of residue substitution within IDRs.

This result is a partial reversal of the typical pattern observed in folded proteins where the generally larger and more hydrophobic structure-promoting residues are subject to strict geometric constraints imposed by the tightly packed hydrophobic core. In contrast, the smaller and more hydrophilic disorder-promoting residues are more variable, as they often occur in flexible, solvent-exposed regions. It is, however, consistent with other analyses of sequence-function relationships in IDRs. For example, several studies demonstrated that acidic activation domains of transcription factors, which are usually disordered, contain clusters of hydrophobic residues interspersed throughout their largely acidic chains~\cite{Ravarani2018, Staller2018, Erijman2020}. Additionally, these studies showed that many distinct sequences can yield similar levels of transcriptional activity. Together, these findings suggest a model of transcriptional activation where the repulsions between acidic residues maintain the bulky hydrophobic residues in accessible conformations that in turn allow the activation domains to bind their targets through non-specific hydrophobic interactions. As these interactions do not require highly complementary interfaces, the observed increase in exchangeability coefficients between hydrophobic residues is consistent with this model and may reflect the prevalence of such ``fuzzy complexes'' in IDR interactions. Furthermore, other studies have shown that different disorder-promoting residues have specific effects on the material properties of condensates formed by phase-separating IDRs. For example, in FUS family proteins glycine residues enhance fluidity whereas glutamine and serine residues promote hardening~\cite{Wang2018}. Even glutamine and asparagine residues, which differ by a single methylene group, can have disparate effects on the conformational preferences of IDRs. While glutamine-rich sequences are conformationally heterogeneous and form toxic aggregates, asparagine-rich sequences instead assemble into benign amyloids, as asparagine's shorter side chain promotes the formation of turns and $\beta$-sheets~\cite{Halfmann2011}. Thus, these observations along with the decreased exchangeability coefficients between disorder-promoting residues suggest that subtle differences in their biochemical properties may inform patterns of residue substitution in IDRs.
